--- 
output: github_document
---

### Interpreting Pfizer's results

The goal of the below is to show plausible point estimates and 95% credible intervals based on the protocol-specified prior for Pfizer. This is informative, both for determining the likely vaccine/placebo split observed (which was not detailed in [Pfizer's press release](https://www.pfizer.com/news/press-release/press-release-detail/pfizer-and-biontech-announce-vaccine-candidate-against)) and for showing the uncertainty (credible intervals) that they are likely to report. 

Because randomization is 1:1 in Pfizer's protocol, VE can be approximated as follows: given N observed study endpoints (in this case, symptomatic COVID-19 disease), VE ~ 1 - v/p, where v is the number of endpoints observed in the vaccine arm and p is the number in the placebo arm. In Pfizer's protocol, they specify a Beta(a, b) prior for theta = (1-VE) / (2-VE), where a = 0.700102 and b = 1. Note that theta = v / (v+p) describes the proportion of the N = v + p endpoints that received the vaccine. We can also do some algebra and find that VE = (1 - 2 theta) / (1 - theta); this is relevant for the computations below. 

Now, Pfizer reported VE of "greater than 90%", which is pretty ambiguous, particularly because their protocol does not specify (or I couldn't find it!) how they will summarize their posterior distribution to obtain a point estimate. Let's assume that they are specifying the posterior mean of VE for their point estimate and a central credible interval for their uncertainty. For the credible interval, I am reporting a central 99.5% interval. Again, this is guessing, but the protocol specifies that interim hypothesis tests are based on posterior probability that VE > 30 given the data being > 0.995, so could make sense to report this level. Then again, they changed their interim monitoring plan based on feedback from FDA, so really, who knows what they'll report. 

We can compute a point estimate and credible interval for VE as follows.^[Below, I'm being lazy and using a numeric approximation to the posterior mean of VE; maybe there's an analytic solution, but I'm too lazy at this moment to figure it out 🙃.]


```{r}
# from protocol pg. 111
# https://pfe-pfizercom-d8-prod.s3.amazonaws.com/2020-11/C4591001_Clinical_Protocol_Nov2020.pdf
a <- 0.700102
b <- 1
ci_level <- 0.995

# from press release
total_cases <- 94

# probably a smarter way to do this
prettify <- function(pt_est, ci){
	paste0(sprintf("%.3f", pt_est), " (", 
	       sprintf("%.3f", ci[1]), ", ", 
	       sprintf("%.3f", ci[2]), ")")
}

# figure out pt est and credible interval
vax_cases <- 1:10
pt_est_ci <- rep(NA, length(vax_cases))
for(v in vax_cases){
	p <- total_cases - v
	# posterior beta dist. parameters
	alpha_v <- a + v
	beta_v <- b + p
	theta_post_samps <- rbeta(5e6, alpha_v, beta_v)
	ve_post_samps <- (1 - 2*theta_post_samps) / (1 - theta_post_samps)
	pt_est <- mean(ve_post_samps)
	ci <- quantile(ve_post_samps, c((1 - ci_level)/2, 1 - (1 - ci_level)/2))
	pt_est_ci[v] <- prettify(pt_est, ci)
}

knitr::kable(cbind(vax_cases, pt_est_ci), row.names = FALSE,
             col.names = c("Vax cases", 
                           paste0("VE (", ci_level*100, "% CI)")))
```

So if this is indeed how Pfizer is reporting, then the likely case split given the language in the press announcement seems to be 7 (or 8) vaccine cases and 87 (or 86) placebo. The credible intervals are also informative that we have high posterior belief under Pfizer's prior that VE is really high.

### Update based on full results

Since interim results above, Pfizer reported results based on 170 COVID-19 disease endpoints with 162:8 placebo:vaccine split. We can compute the posterior distribution for VE based on these results and their protocol-specified prior. 

```{r}
total_cases_full <- 170
v <- 8
p <- total_cases_full - v
# specified in protocol for final analysis
ci_level_full <- 0.986
# posterior beta dist. parameters
alpha_v <- a + v
beta_v <- b + p
theta_post_samps <- rbeta(5e6, alpha_v, beta_v)
ve_post_samps <- (1 - 2*theta_post_samps) / (1 - theta_post_samps)
pt_est <- mean(ve_post_samps)
ci <- quantile(ve_post_samps, c((1 - ci_level)/2, 1 - (1 - ci_level)/2))
```

The results indicate an estimate VE (98.6% credible interval) of `r prettify(pt_est, ci)`.

Here is the full posterior distribution.

```{r}
ve_dat <- data.frame(ve = ve_post_samps)
library(ggplot2)
ggplot(ve_dat, aes(ve)) + 
  geom_density() + theme_bw()
```



